Boolean types and semantic subtyping for Featherweight Java

### 2020-03-22 dissertation/l4proj.tex
finish introduction

%
% put any additional packages here
%

@@

%==============================================================================
%% METADATA
\title{Boolean types and semantic subtyping for Featherweight Java}
\author{Artem Usov}
\date{January 18, 2020}

@@

%==============================================================================
%% ABSTRACT
\begin{abstract}
    Every abstract follows a similar pattern. Motivate; set aims; describe work; explain results.
    \vskip 0.5em
    ``XYZ is bad. This project investigated ABC to determine if it was better. 
    ABC used XXX and YYY to implement ZZZ. This is particularly interesting as XXX and YYY have
    never been used together. It was found that  
    ABC was 20\% better than XYZ, though it caused rabies in half of subjects.''
\end{abstract}

%==============================================================================

% EDUCATION REUSE CONSENT FORM
% If you consent to your project being shown to future students for educational purposes
% then insert your name and the date below to  sign the education use form that appears in the front of the document. 
% You must explicitly give consent if you wish to do so.
% If you sign, your project may be included in the Hall of Fame if it scores particularly highly.
%
% Please note that you are under no obligation to sign 
% this declaration, but doing so would help future students.
%
\def\consentname {Artem Usov} % your full name
\def\consentdate {19 January 2020} % the date you agree
\educationalconsent

%==============================================================================
\tableofcontents

%==============================================================================
%% Notes on formatting
%==============================================================================
% The first page, abstract and table of contents are numbered using Roman numerals and are not
% included in the page count. 
%
% From now on pages are numbered
% using Arabic numerals. Therefore, immediately after the first call to \chapter we need the call
% \pagenumbering{arabic} and this should be called once only in the document. 
%
% Do not alter the bibliography style.
%
% The first Chapter should then be on page 1. You are allowed 40 pages for a 40 credit project and 30 pages for a 
% 20 credit report. This includes everything numbered in Arabic numerals (excluding front matter) up
% to but excluding the appendices and bibliography.
%
% You must not alter text size (it is currently 10pt) or alter margins or spacing.
%
%
%==================================================================================================================================
%
% IMPORTANT
% The chapter headings here are **suggestions**. You don't have to follow this model if
% it doesn't fit your project. Every project should have an introduction and conclusion,
% however. 
%
%==================================================================================================================================
\chapter{Introduction}

% reset page numbering. Don't remove this!
\pagenumbering{arabic}

Why should the reader care about what are you doing and what are you actually doing?
\section{Guidance}

\textbf{Motivate} first, then state the general problem clearly.

\section{Writing guidance}
\subsection{Who is the reader?}

This is the key question for any writing. Your reader:

\begin{itemize}
    \item
          is a trained computer scientist: \emph{don't explain basics}.
    \item
          has limited time: \emph{keep on topic}.
    \item
          has no idea why anyone would want to do this: \emph{motivate clearly}
    \item
          might not know \emph{anything} about your project in particular:
          \emph{explain your project}.
    \item
          but might know precise details and check them: \emph{be precise and
              strive for accuracy.}
    \item
          doesn't know or care about you: \emph{personal discussions are
              irrelevant}.
\end{itemize}

Remember, you will be marked by your supervisor and one or more members
of staff. You might also have your project read by a prize-awarding
committee or possibly a future employer. Bear that in mind.

\subsection{References and style guides}
There are many style guides on good English writing. You don't need to
read these, but they will improve how you write.

\begin{itemize}
    \item
          \emph{How to write a great research paper} \cite{Pey17} (\textbf{recommended}, even
          though you aren't writing a research paper)
    \item
          \emph{How to Write with Style} \cite{Von80}. Short and easy to read. Available online.
    \item
          \emph{Style: The Basics of Clarity and Grace} \cite{Wil09} A very popular modern English 
          style guide.
    \item
          \emph{Politics and the English Language} \cite{Orw68}  A famous essay on effective, clear
          writing in English.
    \item
          \emph{The Elements of Style} \cite{StrWhi07} Outdated, and American, but a classic.
    \item
          \emph{The Sense of Style} \cite{Pin15} Excellent, though quite in-depth.
\end{itemize}

\subsubsection{Citation styles}

\begin{itemize}
    \item If you are referring to a reference as a noun, then cite it as: ``\citet{Orw68} discusses
          the role of language in political thought.''
    \item If you are referring implicitly to references, use: ``There are many good books on writing
          \citep{Orw68, Wil09, Pin15}.''
\end{itemize}

There is a complete guide on good citation practice by Peter Coxhead available here: 
\url{http://www.cs.bham.ac.uk/~pxc/refs/index.html}. 
If you are unsure about how to cite online sources, please see this guide: 
\url{https://student.unsw.edu.au/how-do-i-cite-electronic-sources}.

\subsection{Plagiarism warning}

\begin{highlight_title}{WARNING}

    If you include material from other sources without full and correct attribution, you are 
    commiting plagiarism. The penalties for plagiarism are severe. Quote any included text and cite
    it correctly. Cite all images, figures, etc. clearly in the caption of the figure.
\end{highlight_title}

%==================================================================================================================================
\chapter{Background}
What did other people do, and how is it relevant to what you want to do?

@@

Type systems are by far the most popular and best developed methods in modern software engineering 
to reduce the number of run-time errors in a program. A possible definition of a type system is 
given by \citet{Pierce2002} as:

\textit{A type system is a tractable syntactic method for proving the absence of certain program
    behaviors by classifying phrases according to the kinds of values they compute.}

In particular, the most important detail in the above definition is the emphasis on classifying 
phrases in our language into a particular type based on the kinds of values they hold or compute. 
For example this would allow us to check that the arguments given to an arithmetic operation are 
always numbers. This therefore allows us to validate that a program is absent of certain bad 
run-time behaviour.

However what the above definition is missing that is it defines a type system as solely a syntactic 
method, where it is defined by a formal system of deduction rules, which is by far the most common 
approach for a type system. There does exist an alternative method in the semantic approach. 
\citet{Frisch2002} describe the semantic approach as instead starting with a model of the language 
and all the possible values in the language and define an interpretation of types as subsets of 
values the model.

Such as model comes with several advantages, such as when $t \leq s$ does not hold, then it is
possible to exhibit an element of the model which is in the interpretation of type $t$ and not $s$, 
even in the presence of arrow \textit{(function)} types. This can then be used to show more 
informative error messages to the programmer \citep{Castagna2005}. However due to being a more 
technical, such as it being not trivial to actually define the interpretation of types as subsets of
a model, it has received less attention than syntactic methods.

The first problem we encounter in the semantic approach is that in the semantic approach, $t_{1}$
is a subtype of $t_{2}$ if all the $t_{1}$-values are also $t_{2}$-values. However, in this way, 
subtyping is defined by relying on the notion of well-typed values; hence, we need the typing 
relation to determine typing judgements for values; but the typing rules use the subtyping relation
which we are going to define \citep{Dardha2017}. So, there is a circularity as demonstrated in 
Figure \ref{fig:circ}.

To solve this problem, we follow the framework defined in \citep{Frisch2008}. The general idea of
the framework is that we first extend the types in the language with \textit{Boolean Combinators}: 
union $\lor$, intersection $\land$ and negation $\neg$. We then define an abstract model $\beta$ 
where our types are interpreted according to the set-theoretic meaning of the combinators and can 
define subtyping as set-inclusion and therefore define the typing rules. Separately, if we have a 
notion of well-typed values, we can define another interpretation of types as sets of values. We can
again define a subtyping relation $\nu$. While these may be different relation, if the models are 
chosen carefully such that 
\begin{equation}
    s \leq_{\beta} t \iff s \leq_{\nu} t,
\end{equation}
then these models coincide and this closes the circularity.

\begin{figure}
    \centering
    \includegraphics[width=0.4\linewidth]{images/circularity.PNG}
    \caption{TODO caption, \citep{Castagna2005}}
    \label{fig:circ}
\end{figure}

@@

The syntax of the language is exactly that of Featherweight Java, or FJ \citep{Igarashi1999}, with 
the associated type system only being different. This was chosen on purpose as the full features of 
Java such as concurrency and reflection are orthogonal (\textit{TODO reword}) to the purpose of the 
project while the type system of FJ still exactly corresponds to that of Java. The main key 
simplification in FJ is the omission of assignment. All fields of an object are initialised in the
constructor and not changed afterwards. While this restricts FJ to what is essentially a functional 
fragment of Java, it is still fully computationally complete. The syntax of FJ can be seen in Figure
\ref{fig:syntax} and a sample code fragment can be seen in Figure \ref{fig:class}.

\begin{figure}
    \centering
    \includegraphics[width=0.4\linewidth]{images/syntax.PNG}
    \caption{TODO caption, \citep{Igarashi1999}}
    \label{fig:syntax}
\end{figure}

@@

%==================================================================================================================================
\chapter{Design}
How is this problem to be approached, without reference to specific implementation details? 
\section{Guidance}
Design should cover the abstract design in such a way that someone else might be able to do what you did, but with a different language or library or tool.

>>>>

\input{semmacros}

@@

%===================================================================================================

\title{Boolean types and semantic subtyping for Featherweight Java}
\author{Artem Usov}
\date{January 18, 2020}

@@

%===================================================================================================

\begin{abstract}
    Every abstract follows a similar pattern. Motivate; set aims; describe work; explain results.
    \vskip 0.5em
    ``XYZ is bad. This project investigated ABC to determine if it was better.
    ABC used XXX and YYY to implement ZZZ. This is particularly interesting as XXX and YYY have
    never been used together. It was found that
    ABC was 20\% better than XYZ, though it caused rabies in half of subjects.''
\end{abstract}

%===================================================================================================

\def\consentname {Artem Usov}
\def\consentdate {19 January 2020}
\educationalconsent

%===================================================================================================

\tableofcontents

%===================================================================================================

\chapter{Introduction}

\pagenumbering{arabic}

\section{End goal of programming}

The end goal in all programming projects, is to create a solution to the task at hand that performs exactly as expected by the programmer, and exactly solves the problem that was initially defined.
However, this very rarely actually happens.

On the one hand, this is often because the original problem is not statically defined.
If it is a solution for an external client, their requirements and needs will change over time, or the needs and wants of the users using this project can also change.
This means the project needs to be adapted over time.
On the other hand, and almost always much less evident, is that programs do not perform exactly as the programmer imagines them to work.

These are in general called \emph{bugs} in a program, and have become more and more common as the projects we create have become more complex compared to the early days of computing and the underlying hardware we work on has also become more complex.
Therefore in both academia and industry, there have been great efforts over the years to create tools that allow us to decrease the amount of bugs in our programs and increase our productivity.

\subsection{Static Code Analysis}

One area of great effort has been in the development of static code analysis tools.
Some of these exist integrated into the language itself such as with Spark \citep{Carre1990}.
Spark is a formally defined language based on the Ada \citep{Ada1979} language intended for the development of high integrity software such as flight control systems.
Spark uses contracts and flow analysis to logically prove that programs are correct.
Others exist as separate, well-known tools that exist alongside the language such as the Clang analysis tools for C and C++ \citep{kremenek2008}.

While they are not necessarily needed to write correct programs, these tools have quickly become industry standards for maintaining a level of correctness within codebases and avoiding typical programming errors.

\subsection{Into Static Analysis}

Underlying all of the previously mentioned static code analysis tools, is their use of the \textbf{type system} of the language.
Type systems at their most basic work by assigning types to various constructs of a program so that incompatible types cannot be used.

For example, one common area of error in C code that the Clang tools warn about is implicit conversions where the value of an expression is a different type from the one expected, such as the use of a floating point number where an integer is expected.
The C language will implicitly convert the value to an integer, however this can cause unexpected behaviour for the programmer.
It knows that this conversion needs to happen as the type system dictates that you cannot use any other type other than an integer where an integer is expected.

\section{Varying Type Systems}

The effectiveness of static analysis tools and more generally the \emph{safety} and \emph{expressivity} of a programming language therefore depends on the type system that is used.
Type systems have evolved from the relatively simple systems in languages like C to more complex ones such as in Java where there are custom object types and subtyping relations where types can be substituted for one another.

More modern advances in type systems include the use of linear types in the Cyclone \citep{grossman2002} language that allow us to more naturally and logically define finite state machines.
There exist other new developments in type systems such as Boolean types which are explored in this paper.

\section{Goals of this Project}

In this section, the issue and aims of this paper are presented.

\subsection{Problem Statement}

The typical hierarchical subtyping definitions in Java quite often restrict and complicate the logic that the programmer wants to implement. These problems come up particularly often when working with legacy code that has defines incorrect abstractions in its classes. This therefore leads to less understandable and less maintainable code, that could cause problems in the future.

\subsection{Aims}

The aim of this project is to implement a new programming language that uses a proposed type system consisting of boolean types and a semantic subtyping algorithm. This project aims to transform the mathematical definition of such a type system into an actual solution in software, solving the issues encountered doing so. The language should be able to illustrate the improved solutions to the
common example problems encounter in a language without these features such as Java. Furthermore, the language will be evaluated so see whether these additional language features are intuitive to use by programmers.

\section{Dissertation Outline}

The dissertation is structured into seven chapters as follow:

\begin{itemize}
    \item
          \textbf{Chapter 2} provides background about type systems and Featherweight Java, the language on top of which the new type system is implemented. We also discuss some existing work on semantic type systems and boolean types.
    \item
          \textbf{Chapter 3} TODO
    \item
          \textbf{Chapter 4} TODO
    \item
          \textbf{Chapter 5} TODO
    \item
          \textbf{Chapter 6} contains details about the evaluation method, evaluation results and analysis of results.
    \item
          Finally \textbf{chapter 8} concludes the dissertation with a summary of the important results.
\end{itemize}

%===================================================================================================

\chapter{Background}

@@

In all modern programming languages, a type systems is arguably the main way that incorrect behaviour in a program is reduced.
A possible definition of a type system is given by \citet{Pierce2002} as:

\emph{A type system is a tractable syntactic method for proving the absence of certain program behaviors by classifying phrases according to the kinds of values they compute.}

In particular, the most important detail in the above definition is the emphasis on classifying phrases in our language into a particular type based on the kinds of values they hold or compute.
For example by using a type system, we can check that the arguments given to an arithmetic operation are always numbers.
Therefore we can validate that a specific program is absent of certain bad run-time behaviour such as trying to add a boolean value \emph{true} to a number \emph{2}.

However what the above definition is inaccurate about is that is it defines a type system to be solely a syntactic method, in which the system is defined as a list of formal deduction rules.
The syntactic approach is certainly by far the most common approach for a type system.
However, there does exist an alternative method in the semantic approach.
\citet{Frisch2002} describe the semantic approach as instead starting with a model of the language and all the possible values in the language and defines an interpretation of types as subsets of values the model.

Such as model comes with several advantages.
For example, given two types $s$ and $t$, which represent a certain subset of values in the model, when $s \subseteq t$ does not hold, then it is possible to exhibit an element of the model which is in the interpretation of type $s$ and not $t$.
This can then be used to show more informative error messages to the programmer \citep{Castagna2005}, such as showing which value in $s$ is causing the equality not to hold.
However due to being a more technical approach, such as the non-triviality of defining the interpretation of types as subsets of a model, the semantic method has received less attention than syntactic methods.

The first problem we encounter in the semantic approach is that in such a model, $t_{1}$ is a subtype of $t_{2}$ if all the $t_{1}$-values are also $t_{2}$-values, i.e. the equality $t_{1} \subseteq t_{2}$ holds.
However, in this way, subtyping is defined by relying on the notion of well-typed values; hence, we need the typing relation to determine typing judgements for values; but the typing rules use the subtyping relation defined in \citep{Dardha2017}.
So, there is a circularity in our definition, as demonstrated in Figure \ref{fig:circ}.

To solve this problem, we follow the framework defined by \citet{Frisch2008}.
The general idea of the framework is that we first extend the types in the language with \emph{Boolean Combinators}: union $\lor$, intersection $\land$ and negation $\neg$.
We can then define an abstract model $\beta$ where our types are interpreted according to the set-theoretic meaning of the combinators and can define subtyping as set-inclusion and therefore define the typing rules.
Separately, if we have a notion of well-typed values, we can define another interpretation of types as sets of values.
We can again define a subtyping relation $\nu$.
While these may be different relation, if the models are chosen carefully such that:
\begin{equation*}
    s \leq_{\beta} t \iff s \leq_{\nu} t,
\end{equation*}

then these models coincide and this closes the circularity.

\begin{figure}
    \centering
    \includegraphics[width=0.4\linewidth]{images/circularity.PNG}
    \caption{Circularity in the semantic subtyping definition \citep{Castagna2005}}
    \label{fig:circ}
\end{figure}

@@

The syntax of the JSF is exactly that of Featherweight Java (FJ) \citep{Igarashi1999}, with only the associated type system being different.
This was chosen on purpose as the full features of Java such as concurrency and reflection are orthogonal to the purpose of demonstrating the use of the novel semantic type system.
The main key simplification in FJ from Java is the omission of assignment.
All fields of an object cannot be changed after initialisation and all methods are pure functions
While this restricts FJ to what is essentially a functional fragment of Java, it is still fully computationally complete.

\section{Tools}

The type system of FSJ was built using Antlr to define the grammar and type checking of the language and the

@@

%==================================================================================================================================
\chapter{Design}
How is this problem to be approached, without reference to specific implementation details?

\label{sec:design}
The syntax of types is given by the following grammar \citep{Dardha2013,Dardha2017}:
$$
    \begin{array}{llll}
        \tau   & ::=                                       & \alpha \ |\ \mu
               & \mbox{{\em Type term}}
        \\
        \alpha & ::=                                       & \zero \ |\ \btypes \ |\ [\wt{l:\tau}] \ |\ \alpha \ \bf{and}\ \alpha \ |\ \bf{not}\ \alpha
        \qquad
               & \mbox{{\em Object type} ($\alpha $-type)}
        \\
        \mu    & ::=                                       & \alpha \to \alpha \ |\ \mu \ \bf{and}\ \mu \ |\ \bf{not}\ \mu
               & \mbox{{\em Method type} ($\mu $-type)}
    \end{array}
$$

The syntax of terms is given by the following grammar and is based on the standard syntax of terms in FJ \cite{Igarashi1999,Dardha2013,Dardha2017}:
\begin{align*}
     & \mbox{\textit{Class declaration}}  & L \; ::= \; & \syndecl{C}{C}{\wt{\alpha \ a};\ K; \ \wt{M}\ }                         \\
     & \mbox{\textit{Constructor}}        & K \;::=\;   & C\ (\wt{\alpha\ x})\ \{\ \synsuper(\wt{x});\ \wt{\synthis.a}=\wt{x}; \} \\
     & \mbox{\textit{Method declaration}} & M \; ::= \; & \alpha \ m\ (\alpha \ x)\ \{\ \synret e; \}                             \\
     & \mbox{\textit{Expressions}}        & e \; ::=\;  & x\ |\  c\ |\ e.a\ |\ e.m(e) \ | \ \synnew C(\wt{e})
\end{align*}

\section{Guidance}
Design should cover the abstract design in such a way that someone else might be able to do what you did, but with a different language or library or tool.
